Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-22 paper.Rmd
ready to call this final for proofing

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,@beger:etal:2017]. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

## The Effect of Using Smoothed ROC Curves

@@

Furthermore, if what we mean with "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly that's not what B&S are trying to suggest since this would be a problematic way to "evaluate" a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with any eye towards out-of-sample predictive accuracy [cf. @colaresi:mahmood:2017 for how to avoid this problem while leveraging out-of-sample predictions]. Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories may by extension not be externally valid [for a more detailed treatment of this issue, see @fariss:jones:2018].

@@

% MIKE: is it actually true that infant mortality has mostly not been studied in the context of explaining conflict risk? Goldstone in their article claim that it has, but for me that's news.

>>>>

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015;@beger:etal:2017]. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

@@

Furthermore, if what we mean by the concept "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly that's not what B&S are trying to suggest since this would be a problematic way to "evaluate" a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could in fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with any eye towards out-of-sample predictive accuracy [cf. @colaresi:mahmood:2017 for how to avoid this problem while leveraging out-of-sample predictions]. Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize and whose underlying theories may by extension not be externally valid [for a more detailed treatment of this issue, see @fariss:jones:2018].

@@
